---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Sarah Sussman"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)   
library(here)

# Check working directory
here()

NTL_LTER <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)

# Set date to date format
NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%m/%d/%y")

#2

mytheme <- theme_bw(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") #alternative: legend.position + legend.justification

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature for the month of July does change with depth across all lakes
Ha: Mean lake temperature for the month of July does not change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_LTER_July <- NTL_LTER %>%
  filter(month(sampledate) == 07) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()
  
#5
depth_by_temp <- ggplot(NTL_LTER_July, aes(x = depth, y = temperature_C)) +
  ylim(0, 35) +
  geom_point() +
  mytheme +
  geom_smooth(method="lm", se = FALSE, color = "red") +
  labs(x = "Depth (m)", y = "Temperature (°C)", 
       caption = "Relationship between temperature (°C) and depth.")

print(depth_by_temp)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The distribution of points appears to follow a linear pattern. As depth increases, temperature decreases. This makes sense with what we know (as you get further away from the water's surface (increase depth), water temperature also decreases.) It also appears that once a certain depth is hit (~10m) temperature remains the same (around 5°C).


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
regression_depth_temp <- lm(data = NTL_LTER_July, temperature_C ~ depth)
summary(regression_depth_temp)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model results indicate that depth is a significant factor in water temperature (p < 0.05). The degrees of freedom the result is based on is 9726 (9728 observations - 1). The R-squared value is 0.74. Meaning that depth accounts for ~74% of the variance in temperature. Because the R-squared value is cose to 1, it means there are not many other variables that could account for temperature variance. The coefficients indicate that for every 1m increase in depth, the temperature decreases by 1.95°C.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
July_AIC <- lm(data = NTL_LTER_July, temperature_C ~ depth + year4 + daynum)
step(July_AIC)

#10
July_multiple_regression <- lm(temperature_C ~ depth + year4, + daynum, data = NTL_LTER_July)
summary(July_multiple_regression)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests we use to predict temperature in the multiple regression is depth, year4, and daynum. The starting AIC is 26065.53. When year4 is removed, the AIC increases slightly to 26066. When daynum is removed, the AIC increases to 26148. When depth is removed, the AIC sharply increases to 39189. The suggests that depth has the greatest influence on temperature variance, when day number and year have little influence on temperature. Because the AIC increases when these variables are removed compared to when none of the variabkes are removed, all variables are suggested for use. No, using other variables other than depth does not improve the model as the R-squared value from the multiple regression is 0.72, a decrase from the R-squared value of 0.74 when depth is the only variable.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
NTL_LTER_July_anova <- aov(data = NTL_LTER_July, temperature_C ~ lakename)
summary(NTL_LTER_July_anova)

NTL_LTER_July_anova2 <- lm(data = NTL_LTER_July, temperature_C ~ lakename)
summary(NTL_LTER_July_anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The results of the Anova test find that there are significant differences (p < 0.05) between the lakes and their mean temperatures in the month of July. The results of the linear model show that there are significant differences between each lake's mean temperatures and the initial lake used for the analysis (the first lake listed alphabetically, Central Long Lake), as all p-values are < 0.05.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(data = NTL_LTER_July,
       aes(x = depth, y= temperature_C, color = lakename)) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  ylim(0, 35) +
  mytheme +
  labs(x= "Depth (m)",
       y= "Temperature (°C)",
       caption = "Relationship between temperature (C) and depth (m) among each lake in the month of July.")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(NTL_LTER_July_anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake (0.22) and Ward Lake (0.78) have the same mean temperature, statistically speaking, as Peter Lake. The p-value for Peter Lake - Paul Lake is 0.22, which is less than 0.05. The p-value for Ward Lake - Peter Lake is 0.77, which is less than 0.05. A value less than 0.05 would indicate that the mean temperature of the lakes are significantly different. There are few lake pairs that statistically similar, as it seems that the mean temperature of most lakes are statistically different from one another - for this reason it is difficult to point to one lake that is especially statistically distinct from all the other lakes.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two-sample t-test may be another test that could explore whether they have distinct means. A two-sample t-test is used to test the hypothesis that the mean of two samples is equivalent



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
NTL_LTER_July_two_lakes <- NTL_LTER_July %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

lakes.twosample <- t.test(NTL_LTER_July_two_lakes$temperature_C ~ NTL_LTER_July_two_lakes$lakename)
lakes.twosample
```

>Answer: The test result indicates that their mean July temperature is statistically the same, as p-value > 0.05 (it is 0.26). The test reports that the mean temperature for Crampton Lake is 15.35 and the mean temperature for Ward Lake is 14.45. In part 16, The Tukey test result indicates that Crampton Lake and Ward Lake are statistically similar, as the p-value is well above 0.05 with a value of 0.97. So yes, the results of the t-test match the results of the Tukey test in part 16.
